Computer Assisted Verbal Autopsy: Comparing Large Language Models to Physicians for Assigning Causes to 6939 Deaths in Sierra Leone from 2019-2022

### 2025-06-25 manuscript/manuscript.tex
Add mesh compliant keywords
Revising with feedback from GPT

\keywords{Cause of Death, Physician Coding, Verbal Autopsy, GPT, AI, LLM}

@@

Every year, 41 million people died prematurely from noncommunicable diseases, accounting for 74\% of all deaths globally \citep{worldhealthorganizationNonCommunicableDiseases2019}. While most of these deaths are preventable, effective intervention requires evidence-based resource allocation that targets high-risk populations \citep{benzigerGlobalBurdenDisease2016}. Reliable mortality counts and accurate Cause of Death (COD) data are essential for guiding public health policy and reducing premature mortality \citep{lawnMillionNeonatalDeaths2010,lassiCommunityBasedIntervention2015,liuExcessMortalityPersons2017,ewigCommunityacquiredPneumoniaEmergency2011}. However, in many low-income countries, civil registration and vital statistics systems remain incomplete. Fewer than half of all deaths are registered, and among these, only 8\% have an assigned COD \citep{worldhealthorganizationSCOREHealthData2021}. To address this gap, Verbal Autopsy (VA) has been deployed as a scalable method for collecting mortality data and assigning likely CODs, particularly for deaths that occur outside of healthcare facilities, which account for more than half of all deaths \citep{desavignyIntegratingCommunitybasedVerbal2017,thomasVerbalAutopsyHealth2018,rampatigeSystematicReviewStatistics2014,adairWhoDiesWhere2021}.

@@

GPT-3.5/4 require substantial computational infrastructure for training and inference, making local deployment impractical due to cost and model ownership constraints. Consequently, sensitive data, such as identifiable personal information, must be transmitted to external servers, raising significant privacy concerns. Data submitted via prompts, which include narrative content used for COD assignment, may be collected by service providers (e.g., OpenAI) and potentially misused \citep{taoOpeningPandoraBox2023}. Moreover, there is risk that sensitive information could be exposed or exploited through malicious actors or poorly controlled data handling \citep{khowajaChatGPTNeedsSPADE2024, wuUnveilingSecurityPrivacy2024}. In contrast, InterVA-5 and InSilicoVA can be run entirely on local systems, enabling data to remain under the control of the data owner. This approach reduces dependency on external services and better safeguards data privacy.

@@

This section outlines key limitations of the current study related to the use of GPT models. Section \ref{discuss-limits-icd10} discusses the omission of detailed performance evaluation using ICD-10 codes. Section \ref{discuss-limits-model} addresses the need for further parameter tuning and the evaluation of model consistency and multi-COD assignments.

\subsubsection{ICD-10 Evaluation and Low Sample Sizes}\label{discuss-limits-icd10}

This study evaluated model performance using broad CGHR-10 categories rather than specific ICD-10 codes. In practice, physicians assign more detailed ICD-10 codes, but InterVA-5 and InSilicoVA generate only broader WHO VA codes and cannot assign ICD-10 codes directly, partly due to insufficient sample cases for many specific ICD-10 categories to support reliable modeling. For example, even broad CGHR-10 categories had fewer than 10 cases (e.g., congenital anomalies, suicide), and were excluded from evaluation. While GPT models assigned ICD-10 codes, lower performance can be expected, as even physicians show limited agreement on detailed ICD-10 coding, with only 6,939 (58\%) of 11,920 records in agreement, necessitating the use of broader categories (e.g., CMEA-10 codes) to assess equivalence.

@@

>>>>

\keywords{Cause of Death, Physicians, Computer-Assisted Diagnosis, Artificial Intelligence, Natural Language Processing, Machine Learning, Mortality, Surveillance, Mathematical Models, Global Health}

@@

Every year, 41 million people died prematurely from noncommunicable diseases, accounting for 74\% of all deaths globally \citep{worldhealthorganizationNonCommunicableDiseases2019}. While most of these deaths are preventable, effective intervention requires evidence-based resource allocation that targets high-risk populations \citep{benzigerGlobalBurdenDisease2016}. Reliable mortality counts and accurate Cause of Death (COD) data are essential for guiding public health policy and reducing premature mortality \citep{lawnMillionNeonatalDeaths2010,lassiCommunityBasedIntervention2015,liuExcessMortalityPersons2017,ewigCommunityacquiredPneumoniaEmergency2011}. However, civil registration and vital statistics systems remain incomplete in many low-income countries. Fewer than half of all deaths are registered, and among these, only 8\% have an assigned COD \citep{worldhealthorganizationSCOREHealthData2021}. To address this gap, Verbal Autopsy (VA) has been deployed as a scalable method for collecting mortality data and assigning likely CODs, particularly for deaths that occur outside of healthcare facilities, which account for more than half of all deaths \citep{desavignyIntegratingCommunitybasedVerbal2017,thomasVerbalAutopsyHealth2018,rampatigeSystematicReviewStatistics2014,adairWhoDiesWhere2021}.

@@

GPT-3.5/4 require substantial computational infrastructure for training and inference, making local deployment impractical due to cost and model ownership constraints. Consequently, sensitive data, such as identifiable personal information, must be transmitted to external servers, raising significant privacy concerns. Data submitted via prompts, which include narrative content used for COD assignment, may be collected by service providers (e.g., OpenAI) and potentially misused \citep{taoOpeningPandoraBox2023}. There is risk that sensitive information could be exposed or exploited through malicious actors or poorly controlled data handling \citep{khowajaChatGPTNeedsSPADE2024, wuUnveilingSecurityPrivacy2024}. While jurisdictions, such as the European Union, enforce strict protections under the General Data Protection Regulation (GDPR), most low‑ and middle‑income countries are only beginning to formalize regulatory frameworks for data protection and artificial intelligence governance \citep{intersoftconsultingGeneralDataProtection2018,beckProtectingConfidentialitySecurity2016,kwarkyeWeKnowWhat2025}. In contrast, InterVA-5 and InSilicoVA can be run entirely on local systems, enabling data to remain under the control of the data owner. This approach reduces dependency on external services and better safeguards data privacy.

@@

This section outlines key limitations of the current study related to the use of GPT models. Section \ref{discuss-limits-phy} discusses the use of physician assignment as the reference standard for comparing models. Section \ref{discuss-limits-model} addresses the need for further parameter tuning and the evaluation of model consistency and multi-COD assignments, while Section \ref{discuss-limits-global} notes the importance of more diverse datasets for model evaluation.

\subsubsection{Physician Reference Standard}\label{discuss-limits-phy}

This study evaluated model performance using broad CGHR-10 categories rather than specific ICD-10 codes. In practice, physicians assign more detailed ICD-10 codes, but InterVA-5 and InSilicoVA generate only broader WHO VA codes and cannot assign ICD-10 codes directly, partly due to insufficient sample cases for many specific ICD-10 categories to support reliable modeling. For example, even broad CGHR-10 categories had fewer than 10 cases (e.g., congenital anomalies, suicide), and were excluded from evaluation. While GPT models assigned ICD-10 codes, lower performance can be expected, as even physicians show limited agreement on detailed ICD-10 coding, with only 6,939 (58\%) of 11,920 records in agreement, necessitating the use of broader categories (e.g., CMEA-10 codes) to assess equivalence. Reliance on physician assignment as the reference standard may introduce bias, as physician interpretations may be shaped by local epidemiological knowledge, particularly for more complex cases or ambiguous narratives \citep{leitaoComparisonPhysiciancertifiedVerbal2014}.

@@

\subsubsection{Global Validity}\label{discuss-limits-global}

While this study rigorously compares computer algorithms for COD assignment in Sierra Leone, the extent to which these findings are applied to other geographic or epidemiological contexts remains limited. Variations in local mortality profiles, linguistic expression, health system infrastructure, and culturally specific interpretations of illness shape the content and structure of VA narratives and questionnaires \citep{setelScandalInvisibilityMaking2007,fottrellVerbalAutopsyMethods2010,byassImperfectWorldGlobal2010}. For example, the ways in which symptoms are described, terminology used, and aspects emphasized by respondents differ across languages and cultural settings. Moreover, Sierra Leone is predominantly driven by infectious diseases, such as malaria and respiratory infections, a pattern that contrasts with regions where non-communicable diseases typically constitute the leading CODs in North America and Europe, or where violence and road traffic injuries predominate in parts of Latin America and Asia \citep{murrayMortalityCauseEight1997,mathersGlobalRegionalCauses2009,vosGlobalBurden3692020}. Given ongoing efforts to scale and integrate VA systems for mortality surveillance across diverse low- and middle-income countries, further validation across globally representative VA datasets is essential to evaluate model robustness, adaptability, and operational utility in practice \citep{shawonRoutineMortalitySurveillance2021,maqungoCanVerbalAutopsies2024,onyangoUsingVerbalAutopsy2024}.
